
# Hypothesis Testing and Interval Estimation; Answering Research Questions {#chp4}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, warning = FALSE,
                      options(digits = 3, scipen = 999))
```

## Computing Corner

We will learn the basisc for hypothesis testing in R.

### Probability Distributions in R

For every probability distribution there are four commands.  These command for each distribution are prepended by a letter to indicate the functionality.

- "d" returns the height of the probability "d"ensity function
- "p" returns the cummulative density function or the "p"robability of being being between two values of the random variable.
- "q" returns the inverse density function or the value of the random variable ("q"uantile) given a probability.
- "r" returns a "r"andomly generated number from the probability distribution

The distributions you are most likely to encounter in econometrics are the normal (norm), the F distribution (f), the chi-square distribution (chisq), and Student's t-distribution (t).  Others include the uniform (unif), binomial (binom), Poisson (pois), etc.  Use of the help tab in the Files/Plots/Packages/Help pane or use of `args` will list the arguments necessary to extract value for each distribution.  

### Critical Values in R

To calculate critical values to perform a hypothesis test use the "q" version of the probability distribution.  This will return the quantile for the given probability.  The probability under the curve will be cummulative from $-\infty$ to the quantile returned.  The "q" version will return the critical value for a one-tail test.  Suppose you'd like to test the following hypothesis about $\mu$:

$$H_0:\mu=0$$
$$H_1:\mu<0$$
at the $\alpha=.05$ level of significance.  To calculate the critical t-stastic call `qt(p = .05, df = n-1)`.  You know from `args(qt)` the default value of the argument lower.tail is TRUE.  Suppose, instead, you'd like to test the following hypothesis about $\mu$

$$H_0:\mu=0$$
$$H_1:\mu>0$$
at the $\alpha = .10$ level of significance.  You can call `qt` in two ways:

1. `qt(p = .10, df = n-1, lower.tail = FALSE)` or
2. `qt(p = .90, df = n-1)`

Finally, suppose you'd like to test the following hypothesis about $\mu$

$$H_0:\mu=0$$
$$H_1:\mu\ne0$$
at the $\alpha=.01$ level of significance.  Since the t-distribution is symmetric you can use the lower tail or upper tail value and -1 times it.  You can call `qt` in three ways:

1. `qt(p = .005, df = n-1)` or
2. `qt(p = .005, df = n-1, lower.tail = FALSE)` or
3. `qt(p = .995, df = n-1)`

You can find crtical values for the normal, F, and $\chi^2$ distributions with similar function calls.

#### *p* values in R

To calculate _p_ values in R, use the "p" version of the distribution call.  So suppose we test the following hypothesis:

$$H_0:\sigma_1^2=\sigma_2^2$$
$$H_0:\sigma_1^2\ne\sigma_2^2$$

at the $\alpha=.05$ level of significance.  We could use an F test of the form

$$F=\frac{s_x^2}{s_y^2}$$ 

where $s_x^2$ and $s_y^2$ are the sample variances with n-1 and m-1 degrees of freedom. To calculate the *p* value, call `pf(F, n-1, m-1)` where F is the value calculated above.   

### Confidence Intervals for OLS estimates

In addition to `confint()`, `confint_tidy()` from the broom package will create a tibble of the low and high values for each estimate.  The default level of confidence is 95%. 

### Power Curves

The power curve represents the probability of making Type II error under alternative null hypotheses.  We can generate the power of the test with the `pwr.norm.test(d = NULL, n = NULL, sig.level =.05, power = NULL, alternative = c("two-sided", "less", "greater"))` call from the `pwr` package and plot the power with `ggplot`. To estimate the power we need the effect size $d = \beta_i - \beta$ where $\beta$ is the hypothesised paramater.  We will use 
$$H_0: \beta = 0$$
$$H_1: \beta > 0$$
 
The $\beta_i$ represent alternative null hypothseses for $\beta$. Let's let $0 < beta < 7$. Let the significance level be $\alpha=.01$ and $se_{\beta} = 1$.  

```{r power}
library(tidyverse)
library(pwr)

beta_i <- seq(0, 7, .1)
se_beta <- 1 # to keep se_beta = 1 we will set n = 1 below.

pwr <- pwr.2p.test(beta_i, n = 1, sig.level = .01, alternative = "greater")

#the output is a list we need to extract, h and power from pwr

data <- tibble(beta = pwr$h, power =  pwr$power)

data %>% 
ggplot(aes(x = beta, y = power)) +
  geom_line() +
  ylab("Probability of rejecting the null") +
  xlab(expression(beta))
```


