# Experiments: Dealing with Real--World Challenges {#chp10}


```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, warning = FALSE,
                      options(digits = 3, scipen = 999))
```

We will learn to assess balance with R in this chapter. We need the following libraries

```{r}
library(tidyverse)
library(broom)
```

## Assess Balance

Let's use the `ProgramEffectiveness` data set from the `AER` package to assess balance.  The `ProgramEffectiveness` data set contains 32 observations on four variables^[`?AER::ProgramEffectiveness for more information`]. The data are used to examine whether a new method of teaching economics improved performance in later economics courses.  The variables are *grade* coded as a factor with levels "increase" and "decrease", *average* (grade point average), *testscore* (test score on economics test), and *participation* coded as a factor with levels "no" and "yes".  *participation* is the treatment in this case.  We assess the balance below:

```{r}
library(AER)
data("ProgramEffectiveness")
ProgramEffectiveness %$%
  lm(average ~ participation) %>% 
  tidy()
ProgramEffectiveness %$%
  lm(testscore ~ participation) %>% 
  tidy()
```

For each variable, we can conclude that the treatment is balanced.  

## Estimate ITT Model

We estimate the ITT model below:

```{r}
ProgramEffectiveness %$%
  lm(as.numeric(grade) ~ participation) %>% 
  summary()
```

We can reject the null hypothesis of no effect and conclude that participation increased the test score on later tests.

