
# Regression Discontinuity: Looking for Jumps in Data {#chp11}

```{r 11-setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, warning = FALSE,
                      options(scipen = 999, digits = 3))
```

We will learn techniques in R to deal with "jumps" in the data. We will use the following libraries

```{r 11-packages}
library(tidyverse)
```

## Same slope

To estimate an RD model where the slope is the same before and after the cutoff value make use of the `ifelse` call in R.  `ifelse` returns one value if the test condition holds and another when it doesn't.  For example suppose the we create a variable, *T* that takes on the value 1 when another variable say *X* is greater than 10.  Create *T* with the call `T -> ifelse(X > 10, 1, 0)`^[Using *T* as variable name is not good practice as T is an abbreviation for TRUE, so we will use *D* throughout our code.]. 

Let's estimate an RD model using the data from a 2009 paper by Carpenter and Dobkin about the effect of increasing the drinking age on mortality rates.^[Carpenter, Christopher and Carlos Dobkin. "The Effects of Alcohol Consumption on Mortality: Regression Discontinuity from the Minimum Drinking Age," *American Economic Journal: Applied Econometrics*, 2009, 1:1, 164-182.  
The data are available at https://github.com/jrnold/masteringmetrics/tree/master/masteringmetrics/data in mlda.rda]. Let's just look at motor vehicle deaths as a function of age.

```{r 11-1}
load("Data/mlda.rda")
mlda %>% 
  ggplot(aes(x = agecell, y = mva)) + 
  geom_point(na.rm = TRUE) +
  geom_vline(xintercept = 21) + 
  labs(y = "Deaths in Moving Vehicle Accidents", x = "Age")
```

There appears to be a discontinuity at age 21.  Let's estimate the RD model $$mva = \beta_0+\beta_1T+\beta_2(agecell-21)+\epsilon$$  
where  
$$\begin{aligned}
T &= 1 \text{ if agecell} \geq 21 \\
T &= 0 \text{ if agecell} < 21
\end{aligned}$$

We will make use of the `tidyverse` verb `mutate` and pipe operators to create *D*^[Recall we will use *D* to avoid the ambiguity of *T* as a variable name.].  

```{r 12-constant-slope}
mlda %>% 
  mutate(D = ifelse(agecell >= 21, 1, 0)) %>% 
  lm(mva ~ D + I(agecell - 21), .) %>% 
  jtools::summ()
```

## Varying Slopes 

Let's estimate the relationship described above with a varying slopes RD model.  The model now has the form:  $$mva = \beta_0+\beta_1T+\beta_2(agecell-21)+\beta_3(agecell-21)T+\epsilon$$  
where  
$$T = 1 \text{ if agecell}\geq21$$ $$T = 0 \text{ if agecell}<21$$
```{r 12-varying-slopes}
mlda %>% 
  mutate(D = ifelse(agecell >= 21, 1, 0)) %>% 
  lm(mva ~ D * I(agecell - 21), .) %>% 
  jtools::summ()
```

## Plot RD Model

Use `ggplot` to plot the RD model. We include plots with an simple regression and an RD model.

```{r 12-plots, message=FALSE}
mlda %>% 
  select(agecell, mva) %>% 
  mutate(D = as.factor(ifelse(agecell >= 21, 1, 0))) %>% 
  ggplot(aes(x = agecell, y = mva)) +
  geom_point(aes(color = D), na.rm = TRUE) + 
  geom_smooth(method = "lm", na.rm = TRUE)

mlda %>% 
  select(agecell, mva) %>% 
  mutate(D = as.factor(ifelse(agecell >= 21, 1, 0))) %>% 
  ggplot(aes(x = agecell, y = mva, color = D)) +
  geom_point(na.rm = TRUE) + 
  geom_smooth(method = "lm", na.rm = TRUE)
```


## `rddtools` package

We can estimate RD models with the `rddtools` package^[`?rddtools` for more.].  To estimate an RD model with `rddtools` first create an `rdd_data` object as follows: `rdd_data(y = df$y, x = df$x, cutpoint = C)`.  Use the `rdd_data` object with `rdd_reg_lm` to estimate the model. 

### Same slope 

To estimate an RD model with a constant slope call the argument `rdd_reg_lm(rdd_object, slope = "same")`


```{r rddtools}
rddtools::rdd_data(mlda$mva, mlda$agecell, cutpoint = 21) %>% 
  rddtools::rdd_reg_lm(slope = "same") %>% 
  jtools::summ()
```

Note the results are the same as above.

### Varying Slopes

To estimate an RD model with varying slopes, change the slope argument to "separate".

```{r rddtools-varying}
rddtools::rdd_data(mlda$mva, mlda$agecell, cutpoint = 21) %>% 
  rddtools::rdd_reg_lm(slope = "separate") %>% 
  jtools::summ()
```

### Scatter Plot

```{r}
rddtools::rdd_data(mlda$mva, mlda$agecell, cutpoint = 21) %>% 
  rddtools::rdd_reg_lm(slope = "same") %>% 
  plot()
rddtools::rdd_data(mlda$mva, mlda$agecell, cutpoint = 21) %>% 
  rddtools::rdd_reg_lm(slope = "separate") %>% 
  plot()
```

