# Time Series: Dealing with Stickiness over Time {#chp13}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, warning = FALSE,
                      options(digits = 3, scipen = 999))
```

In this chapter we will learn to work with time--series data in R.

## Time Series Objects in R

Working with time-series data in R is simplified if the data are structured as a *time-series object*.  There are a host of functions and packages dedicated to time-series data. A time-series object is an R structure that contains the observations, the start and end date of the series, and information about the frequency or periodicity.

We will use the [Bike Sharing Dataset](https://archive.ics.uci.edu/ml/machine-learning-databases/00275/)^[Fanaee-T, Hadi, and Gama, Joao, 'Event labeling combining ensemble detectors and background knowledge', Progress in Artificial Intelligence (2013): pp. 1-15, Springer Berlin Heidelberg] from the UCI Machine Learning Repository. 
  
  - The data set has 731 observations on 17 variables
	- instant: record index
	- dteday : date
	- season : season (1:spring, 2:summer, 3:fall, 4:winter)
	- yr : year (0: 2011, 1:2012)
	- mnth : month ( 1 to 12)
	- hr : hour (0 to 23)
	- holiday : weather day is holiday or not (extracted from http://dchr.dc.gov/page/holiday-schedule)
	- weekday : day of the week
	- workingday : if day is neither weekend nor holiday is 1, otherwise is 0.
	+ weathersit : 
		- 1: Clear, Few clouds, Partly cloudy, Partly cloudy
		- 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
		- 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
		- 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog
	- temp : Normalized temperature in Celsius. The values are divided to 41 (max)
	- atemp: Normalized feeling temperature in Celsius. The values are divided to 50 (max)
	- hum: Normalized humidity. The values are divided to 100 (max)
	- windspeed: Normalized wind speed. The values are divided to 67 (max)
	- casual: count of casual users
	- registered: count of registered users
	- cnt: count of total rental bikes including both casual and registered
	
```{r}
library(tidyverse)
library(broom)
library(xts)
library(magrittr)
bike_share <- read_csv("Data/day.csv")
head(bike_share)
```

We can convert `bike_share` to a time-series object using the `xts` package. A time-series object requires an index to identify each observation by its date. We create the index with `seq(as.date("YYYY-MM-DD"), by = "period", length.out = n)`, our start date is 2011-01-01, by "days", with the number of observations as the length.  After we create the date index, we will use `dplyr:select` to choose the variables (we don't need *dteday*, *yr*, or *mnth*). As always, we will make use of the pipe operator to complete the code. 

```{r}
dates <- seq(as.Date("2011-01-01"), by = "days", length.out = 731)
bike_ts <- 
bike_share %>% 
  select(instant, season, holiday, weekday, workingday, weathersit, temp, atemp, hum, windspeed, casual, registered, cnt) %>% 
  xts(dates) 
bike_ts %>% 
  head()
```

We see that the data now have a date index indicating to which date each observations belongs.

## Detecting Autocorrelation

Let's estimate the total number of riders as a function of time, $cnt=\beta_0+\beta_1instant+\epsilon$, and test the residuals for first order auto correlation using the auxiliary regression approach.

```{r}
bike_ts %$%
  lm(cnt ~ instant) 
# retrieve the residuals as e
e <- 
  bike_ts %$%
  lm(cnt ~ instant)$residuals
# auxiliary regression
lm(e ~ lag(e,1)) %>% 
  tidy()
```

The *t-statistic* on $\hat\rho_{t-1}$ is 30.50 so we can reject the null hypothesis of no autocorrelation in the error term.

## Correcting Autocorrelation

### Newey-West

The `sandwich` package contains a function to estimate Newey-West standard errors.^[Unfortunately the pipe operator does not play nice with `NeweyWest`.]  The output of the function call is the corrected variance-covariance matrix.  We still need to calculate *t-statistics* based on the corrected variance-covariances.  We will use the `lmtest` package to perform this test.  `coeftest(lm_object, vcov = variance-covariance_matrix)`.  

```{r}
# estimate the model (the lm_object)
bike_lm <- lm(cnt ~ instant, bike_ts)
bike_lm %>% 
  tidy()
# determine the number of lags for the Newey-West correction
lags <- length(bike_ts$instant)^(.25)
nw_vcov <- sandwich::NeweyWest(bike_lm, lag = lags, prewhite = F, adjust = T)
# model with corrected errors
lmtest::coeftest(bike_lm, vcov. = nw_vcov)
```

### Cochrane-Orcutt 

The `orcutt` package allows us to use the Cochrane-Orcutt method to $\rho$ difference the data to produce corrected standard errors using `cochrane.orcutt(lm_oject)`

```{r}
bike_lm %>%
  orcutt::cochrane.orcutt() %>% 
  tidy()
```

## Dynamic Models

Using a time-series object makes running dynamic models as easy as calling the argument `lag(variable_name, number_of_lags)`.  Suppose we'd like to estimate the a lagged version of the model we have been using with the form $cnt_t=\beta_0+\beta_1instant_t+\beta_2temp_{t-1}+\epsilon$.  We want to see if yesterday's weather affects today's rentals.

```{r}
bike_lm_dyn <- lm(cnt ~ instant + lag(temp, 1), bike_ts)
bike_lm_dyn %>% 
 tidy
```

## Dickey-Fuller Test

The `tseries` package includes an augmented Dickey-Fuller test, `adf.test(time_series)`.

```{r}
bike_ts$cnt %>% 
  tseries::adf.test()
```

We conclude that *cnt* is non-stationary.  

## First Differencing

A simple solution to non-stationarity is to use first differences of values, i.e., $\Delta y_t=y_t-y_{t-1}$.  `diff(x, ...)` makes this easy with a time-series object.  Let's test $\Delta y_t$ for stationarity.

```{r}
bike_ts$cnt %>% 
  diff() %>% 
  tseries::na.remove() %>% # first differencing introduces NA's into the data
  tseries::adf.test()
```

We can reject the null-hypothesis of non-stationarity.  So let's estimate the model $\Delta cnt_t=\beta_0+\beta_1\Delta temp_t+\eta_t$

```{r}
lm(diff(cnt) ~ diff(temp), bike_ts) %>% 
  tidy()
# Compare to same equation in the levels.
lm(cnt ~ temp, bike_ts) %>% 
  tidy()
```

