# Advanced Panel Data {#chp15}

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, warning = TRUE, error = FALSE,
                      options(digits = 3, scipen = 999))
```

In this chapter we will learn techinques in R for panel data where there might be serially correlated errors, temporal dependence with a lagged dependent variable, and random effects models.

## The Data

We will make use of the `Cigar` dataset from the `plm` package for this chapter. `Cigar` is a panel of 46 U.S. states over the period 1963-1992.  The variables are:

* state - State number
* year 
* price - the price per pack of cigarettes in cents
* pop - state population in thousands
* pop16 - state population over the age of 16 in thousands
* cpi - consumer price index (1983=100)
* ndi - per capita disposable income in dollars
* sales - cigarette sales per capita in packs
* pimin - minimum price in adjoining states per pack of cigarettes in cents

```{r}
library(plm)
data("Cigar")
```


The `plm` packages offers many functions to simplify the handling of advanced panel data.

## Variation within Units

`dplyr` verbs make checking for variation within units across multiple variables relatively simple. First we use `group-by` so that any functions will be applied to each state individually.  `summarize_all` will apply a function to each variable. 

```{r}
library(tidyverse)
library(broom)
library(magrittr)
# Check for variaton by state.
Cigar %>% 
  group_by(state) %>% # ensures that subsequent functions will be performed by state
  select(price, pop, pop16, cpi, ndi, sales, pimin) %>% 
  summarise_all(sd) # sd is standard deviation 
# Check for variation by year.
Cigar %>% 
  group_by(year) %>% 
  select(-year, -state) %>% # the "-" indicates variables to be removed
  summarise_all(sd)
```

CPI is the only variable with a standard deviation of 0 for all units.  As would be expected, CPI should not vary within year.  

`pvar` from the `plm` package will perform the task of checking for variation.

```{r}
pvar(Cigar)
```

## Two-Way Fixed Effects Model

Let's estimate cigarette demand as: $$sales_{it}=\beta_0+\beta_1price_{it}+\beta_2pop16_{it}+\beta_3ndi_{it}+\alpha_i+\tau_t+\nu_{it}$$

We would expect $\beta_1<0$, $\beta_2>0$, and $\beta_3<0$ if cigarettes are an inferior good^[If cigarettes are a normal good we'd expect $\beta_3>0$]. 

```{r}
cigar_plm <- plm(sales ~ price + pop16 + ndi, 
                 data = Cigar, # recall plm does not play nice with the expose pipe, %$%
                 index = c("state", "year"), 
                 model = "within", 
                 effect = "twoways")
cigar_plm %>% 
  tidy()
```

Each of the coefficients has the expected sign and is significant at the 5% level.

## Testing for autocorrelation

Testing for autocorrelation is done by testing the following hypothesis: $$H_0:\rho=0$$ $$H_1:\rho\ne0$$

```{r}
Cigar %>% 
  glimpse()
```

Our data are organized by unit by year, so we can estimate $\hat\rho$ directly.  First, obtain the residuals, *e*, from the estimated equation.  Estimate the equation $e=\rho e_{i,t-1}+\eta_{it}$.

```{r}
# Obtain the residuals
Cigar$e <- cigar_plm$residuals
# test of rho hat
aux_1 <- 
  Cigar %$%
  lm(e ~ -1 + lag(e)) # -1 removes the constant.
aux_1 %>% 
  tidy()
```

We can reject the null hypothesis at the 1% level.

We can also check for autocorrelation with the LM test by estimating the model $$\hat\epsilon_{it}=\rho\hat\epsilon_{i,t-1}+\gamma_1price_{it}+\gamma_2pop16_{it}+\gamma_3ndi_{it}+\eta_{it}$$ where $nR^2\sim\chi^2_{df=1}$.

```{r}
aux_2 <- 
plm(e ~ lag(e) + price + pop16 + ndi,  
    data = Cigar,
    index = c("state", "year"), 
    model = "within", 
    effect = "twoways") 
nR2 <- 
  aux_2 %>% 
  r.squared *
  aux_2$df.residual 
nR2 %>% 
  pchisq(1, lower.tail = F)
```

Again, we can reject the null hypothesis of no autocorrelation.

`pwartest` from the `lpm` package allows us to test for autocorrelation (`?pwartest` for relevant arguments).  

```{r}
pwartest(cigar_plm)
```

We reject the null hpothesis of no autocorrelation.

## Estimating $\hat\rho$

To correct for autocorrelation we need an estimate of $\hat\rho$. We can estimate $\hat\rho$ using either auxiliary regression from above.

```{r}
aux_1 %>%  
   tidy()
```

Our estimate of $\hat\rho$ is `0.888` is `r aux_1$coef[1]`.  

```{r}
aux_2 %>% 
  tidy()
```

Our estimate of $\hat\rho$ is `r aux_2$coef[1]`. 

## Estimate a $\rho$-Transformed Model

We can manually transform the data and compare the transformed model to the non-transformed model.

```{r}
rho_hat <- aux_2$coefficients[1] # set rho_hat to the coef of lagged e in aux_2
plm(I(sales - rho_hat*lag(sales)) ~ 
      I(price - rho_hat*lag(price)) + 
      I(pop - rho_hat*lag(pop16)) + 
      I(ndi - rho_hat*lag(ndi)), 
    data = Cigar, 
    index = c("state", "year"), 
    model = "within", 
    effect = "twoways") %>% 
  summary()
cigar_plm %>% 
  tidy()
```

Now only $\hat\beta_1$ is significantly different than zero.

We can use the `panelAR` package to directly estimate a corrected model^[Note the slight differences, because `panelAR` also corrects for heteroscedasticity.]. `?panelAR` for arguments necessary to estimate the corrected model.

```{r}
library(panelAR)
panelAR(sales ~ price + pop +ndi, 
        data = Cigar, 
        panelVar = "state", 
        timeVar = "year",
        autoCorr = "ar1",
        panelCorrMethod = "pcse") %>% 
  summary()
```

## Lagged Dependent Variable Panel Data Model

Let's estimate the lagged-depdendent variable model $$sales_{it} = \gamma sales_{i,t-1}+\beta_0+\beta_1price_{it}+\beta_2pop16_{it}+\beta_3ndi_{it}+\epsilon_{it}$$

```{r}
cigar_lag_plm <- plm(sales ~ lag(sales) + price + pop16 + ndi, 
                 data = Cigar, # recall plm does not play nice with the expose pipe, %$%
                 index = c("state", "year"), 
                 model = "within", 
                 effect = "twoways")
cigar_lag_plm %>% 
  summary()
```

## Random Effects Model

```{r}
plm(sales ~ price + pop16 + ndi,
    data = Cigar,
    index = c("state", "year"),
    model = "random",
    effect = "twoways" 
    ) %>% 
  summary()
```

